---
title: "SmokingNA-ABMQ"
output: html_document
date: '2024-01-29'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```



### Nick's data - Positive autobiographical memories in cigarette smokers. 
- cigarette smokers recalled positive memories in an interview format, for each memory they rated:
    - how long ago the event occurred (ordinal, higher values are more recent memories)  
    - if the memory involved cigarette smoking  
    - how frequently they think about the memory  
    - how (negative-positive) they feel while recalling the memory  
    - how vivid the memory is  

```{r read-data}
library("tidyverse")
library("lme4")
mem_init_tib <- read_csv("data/mem_initial_ratings.csv") |> mutate(id = as.character(id))
```

## Hypothesis: In cigarette smokers, the vividness of a memory is related to (a) how old it is, (b) if it involves smoking, and (c) if it is frequently recalled. 

#### First let's look at the vividness ratings (the DV)  
```{r quick-look-vivid}
quants <- quantile(mem_init_tib$mem_vividness, c(.25,.5,.75), na.rm = TRUE)
vividness_mean <- mean(mem_init_tib$mem_vividness, na.rm = TRUE)
p1 <- mem_init_tib %>% drop_na(mem_vividness) %>%
  ggplot( aes(x=mem_vividness)) + geom_histogram(binwidth=1) + 
    geom_vline(xintercept = quants, colour="red", linetype = "dashed") +
    geom_vline(xintercept = vividness_mean, colour="black") +
    theme_classic() + labs (title = "Mem vividness distribution- all individuals")
p2 <- mem_init_tib %>% drop_na(mem_vividness) %>%
    ggplot( aes(sample=mem_vividness)) + geom_qq() + geom_qq_line() + theme_classic() +
      labs (title = "Mem vividness Q-Q- all individuals")
p1; p2
```
#### Now the time variable (how old is the memory?)  
```{r quick-look-time}
quants <- quantile(mem_init_tib$mem_time, c(.25,.5,.75), na.rm = TRUE)
time_mean <- mean(mem_init_tib$mem_time, na.rm = TRUE)
p1 <- mem_init_tib %>% drop_na(mem_time) %>%
  ggplot( aes(x=mem_time)) + geom_histogram(binwidth=1) + 
    geom_vline(xintercept = quants, colour="red", linetype = "dashed") +
    geom_vline(xintercept = time_mean, colour="black") +
    theme_classic() + labs (title = "Mem time distribution- all individuals")
p2 <- mem_init_tib %>% drop_na(mem_time) %>%
    ggplot( aes(sample=mem_time)) + geom_qq() + geom_qq_line() + theme_classic() +
      labs (title = "Mem time Q-Q- all individuals")
p1; p2

```

#### Now the frequency variable (how often do you think about the memory?)  
```{r quick-look-freq}
quants <- quantile(mem_init_tib$mem_frequency, c(.25,.5,.75), na.rm = TRUE)
frequency_mean <- mean(mem_init_tib$mem_frequency, na.rm = TRUE)
p1 <- mem_init_tib %>% drop_na(mem_frequency) %>%
  ggplot( aes(x=mem_frequency)) + geom_histogram(binwidth=1) + 
    geom_vline(xintercept = quants, colour="red", linetype = "dashed") +
    geom_vline(xintercept = frequency_mean, colour="black") +
    theme_classic() + labs (title = "Mem frequency distribution- all individuals")
p2 <- mem_init_tib %>% drop_na(mem_frequency) %>%
    ggplot( aes(sample=mem_frequency)) + geom_qq() + geom_qq_line() + theme_classic() +
      labs (title = "Mem frequency Q-Q- all individuals")
p1; p2

```


# Fit the full model: vividness explained by time, smoking, and frequency
- we get a singular fit warning - therefore, the model fit results are not interpretable.  
- in linear mixed models, a singular fit occurs when random effect variance is estimated very near zero or collinearity between random effects is very near 1 or -1. This can occur when the data is not sufficiently informative   
- in this case the `mem_frequency` random effect (by-subject slope) is highly negatively correlated with the by-subject intercept (possibly due to lack of variance in the mem_frequency measure)   
- we can examine the `mem_frequency` variable more closely to see this (Q: how did we know that this might be the problem? A: During data collection we suspected that some participants did not understand this measure fully)  
```{r fit-full-model}
model1   = lmer(mem_vividness ~ mem_time + mem_smoking + mem_frequency + (mem_time + mem_smoking + mem_frequency| id), data=mem_init_tib)

summary(model1)

#plot individual frequency ratings by subject
#mem_init_tib |> ggplot(aes(x = id, y = mem_frequency)) +
#  geom_jitter(shape = 1, width = .1, height = 0)

```

## so we will delete frequency from the model  

```{r freq-removed}
model1   = lmer(mem_vividness ~ mem_time + mem_smoking  + (mem_time + mem_smoking | id), data=mem_init_tib, control = lmerControl(optimizer = "bobyqa"))

summary(model1)

```
## smoking is not informative, but time explains vividness  
- recent memories are more vivid
- to do a significance test, let's compare a model with `mem_time` to a model with that fixed effect removed (using R's `anova()` function)  

```{r}
d <- mem_init_tib
baseline  <- lmer(mem_vividness ~ 1 + (mem_time | id), data=d, REML = FALSE)
model1   = lmer(mem_vividness ~ mem_time  + (mem_time | id), data=d, control = lmerControl(optimizer = "bobyqa"))

summary(model1)
anova(model1,baseline)
```

## Extra - mixed effects logistic model  
- does the age of a memory explain whether it is smoking or not?  
- see model below - more recent memories are likely to involve smoking  
```{r}
d <- mem_init_tib |> filter(mem_smoking<3) 
d <- d |> mutate(mem_smoking = 2 - mem_smoking)
smoke_model  <- glmer(
  mem_smoking ~ mem_time + (mem_time |id), 
  family = "binomial", data=d)
summary(smoke_model)
smoke_baseline <- glmer(
  mem_smoking ~ 1 + (mem_time |id), 
  family = "binomial", data=d)
summary(smoke_baseline)
anova(smoke_model,smoke_baseline)
performance::r2(smoke_model)

```

